---
title: 'Final Project'
author: Tom Cairns
output: html_document
---

# Business Understanding

For my final project I am creating a model to classify type of breast cancer from
gene expression data. I am using data taken from Kaggle at:
https://www.kaggle.com/datasets/raghadalharbi/breast-cancer-gene-expression-profiles-metabric/download'


# Data Understanding

```{r import-libraries}
library(tidyverse)
library(caret)
library(class)
library(gmodels)
library(vcd)
library(nnet)
library(imbalance)
```

## Explore Data

```{r}
# Import the data
# bc_filepath <- 'https://www.kaggle.com/datasets/raghadalharbi/breast-cancer-gene-expression-profiles-metabric/download'
# bc <- read.csv(bc_filepath)
bc <- read.csv('METABRIC_RNA_Mutation.csv')
dim(bc)
```

```{r select-columns}
# Feature selection
bc <- bc %>%
    select(cancer_type_detailed, !contains('mut'))
bc <- bc[,-c(2:30)]
```

In this code block I selected the features of interest. These include the
cancer_type_detailed, which is our target feature, and the gene expression data.

```{r explore-data}
# Create table of target feature
table(bc$cancer_type_detailed)

# Function to check normality
isNormal <- function(col) {
    p <- shapiro.test(col)$p.value
    
    # Check if significant
    alpha <- 0.05
    return(p < 0.05)
}

# Find the normality of each column
normal_cols <- lapply(bc[,-c(1)], isNormal)

# Identify the non normal columns
which(normal_cols == F)

# Remove non-normal columns
bc <- bc[, -c(which(normal_cols == F))]
```

Here we can see that there were 13 features with non normal distributions. Since
I have many features and the algorithms I will be using assume normality in the
data, I removed these columns.

```{r check-colinearity}
# Create correlation matrix
cor.matrix <- cor(bc[,-c(1)])

# Identify columns with potential colinearity
threshold <- 0.9
findCorrelation(cor.matrix, cutoff = threshold)
```

In this section I looked for instances of potential collinearity. I used a threshold
of 0.9 and found no columns with a Pearson's correlation above this value. This
means I can continue my analysis without concern.


# Data Preparation

```{r remove-classes}
# Remove highly imbalanced classes
bc <- bc[bc$cancer_type_detailed %in% c('Breast Invasive Ductal Carcinoma',
                                        'Breast Invasive Lobular Carcinoma'),]

table(bc$cancer_type_detailed)
```

In this section I removed three of the total classes from the column 
`cancer_type_detailed`. I removed these classes since they all had fewer than
25 total members. This would create a high bias in the algorithms which might
ignore these classes. There is still some imbalance in the membership of the
classes, but this is much closer to an acceptable split.

## Split Data
**THIS SHOULD BE MOVED UP**
```{r split-data}
# Split the data to maintain proportions of classes
sample <- createDataPartition(y = bc$cancer_type_detailed,
                              p = 0.7,
                              list = F)

# Split into training and testing data
bc_train <- bc[sample,]
bc_train$cancer_type_detailed <- bc[sample, c('cancer_type_detailed')]
bc_test <- bc[-sample,]
bc_test$cancer_type_detailed <- bc[-sample, c('cancer_type_detailed')]

# Check that the data was split properly
cat('\nProportions of training data classes:')
proportions(table(bc_train$cancer_type_detailed))
cat('\nProportion of testing data classes:')
proportions(table(bc_test$cancer_type_detailed))
```

In this section I split the data into training and validation data sets. I used
the function `createDataPartitions()` to do this after setting the random number
generator and seed. I decided to use a 70/30 split for this project. 

I then checked the proportions of each class to ensure that there were equivalent
proportions in each data set. As we can see, the function worked correctly and
we do have similar proportions for each of the classes in the training and 
validation data. 

## Feature Elimination

```{r}
bc_train$cancer_type_detailed_numeric <- as.numeric(factor(bc_train$cancer_type_detailed))

# Create initial multiple logistic regression model
bc_train.reg <- glm(cancer_type_detailed_numeric ~ ., data = bc_train[,-c(1)], family = 'gaussian')
```

```{r}
# Perform the stepwise backward feature elimination using AIC
bc_train.step <- step(bc_train.reg, direction = 'backward')
```

Here we applied the backward stepwise feature elimination using AIC. The
regression model initially had an AIC of over 2800 and it dropped to 2359.72.
The number of features decreased from 477 to 110.

```{r feature-selection}
# Select the features from the step model
cols.select <- colnames(bc_train.step$qr$qr)

# Remove the intercept column
cols.select <- cols.select[-1]

# Filter dataset to only include these columns
bc_train.filter <- bc_train[, cols.select]
bc_test.filter <- bc_test[, cols.select]
bc_train.filter$cancer_type_detailed <- bc_train$cancer_type_detailed
bc_test.filter$cancer_type_detailed <- bc_test$cancer_type_detailed
```

In this section I created a new dataframe containing only the features selected
for in the multiple logistic regression model. This reduced the number of features
from 477 to 91.

## Oversampling

In this data set I have a major class imbalance problem. This is common in 
gene expression analysis (Ang et al., 2016; Blagus & Lusa, 2012; Mahmudah et al.,
2021; Tabares-Soto et al., 2020). Based on the results by Blagus & Lusa (2012),
it seems that the best method of dealing with this is by using the Synthetic
Minority Oversamplng Technique (SMOTE).

```{r over-sampling}
# Run an oversampling method
mwmote(bc_train.filter, numInstances = 2100, 
       classAttr = bc_train.filter$cancer_type_detailed)

# Simple oversampling
bilc <- filter(bc_train.filter, cancer_type_detailed == 'Breast Invasive Lobular Carcinoma')

# Add to the dataset
bc_train.oversampled <- rbind(bc_train.filter, bilc, bilc, bilc, bilc, bilc, bilc, 
                              bilc, bilc, bilc, bilc)

# Check that it worked
table(bc_train.oversampled$cancer_type_detailed)
```

In this code block I created a simple oversampling method of the data to fix
the class imbalance problem. There were 1050 classes of Breast Invasive Ductal
Carcinoma and 100 instances of Breast Invasive Lobular Carcinoma. I used 
oversampling to increase
the samples of the minority class to be about the same as the majority class.
I needed to make 10 copies of the Breast Invasive Lobulr Carcinoma.

I created the variable `bc_train.oversampled` which contains almost equal
sizes of each class.

# Modeling 

## kNN

```{r}
# Create kNN model
bc.knn <- knn(dplyr::select(bc_train.oversampled, -cancer_type_detailed), 
              dplyr::select(bc_test.filter, -cancer_type_detailed), 
              cl = bc_train.oversampled$cancer_type_detailed, k = 2)

# Create Cross Table to verify accuracy
knn.cross <- CrossTable(x = bc_test.filter$cancer_type_detailed, y = bc.knn, 
                        prop.chisq = F)
```

In this section I created a k nearest neighbors model using 3 values of k. 
This model has an accuracy of 74.8%. 

The precision of Breast Invasive Ductal Carcinoma is 83.6% (399 / 477) and the
recall is 86.67% (399 / 450).

The precision of Breast Invasic Lobular Carcinoma is 22.2% (8 / 31) and the 
recall is 19% (8 / 42).

The precision of Breast Mixed Ductal and Lobular Carcinoma is 17.4% (8 / 46)
and the recall is 12.9% (8 / 62).

From this model we can see that the most accurate classification is Breast
Invasive Ductal Carcinoma. Neither of the other two categories are well
classified. I think this is largely due to the imbalance in the categories
of the data set.

```{r}
# Function to find the Accuracy
findAccuracy <- function(cm) {
    accuracy = (cm[1] + cm[4]) / sum(cm)
}

# Find the Accuracy
accuracy.knn <- findAccuracy(knn.cross$t)
paste('The accuracy of this kNN model is:', round(accuracy.knn, 3))
kappa.knn <- Kappa(knn.cross$t)
paste('The kappa statistic of the kNN model is:', 
      round(kappa.knn$Unweighted[[1]], 3))
```

This model had an overall accuracy of 74.2% but a Cohen's kappa statistic of
only 0.092. This shows poor performance by the model and reinforces the concern
that the imbalance in the data set is introducing bias in the model.


## SVM

```{r}
# Create the model
bc_svm <- train(cancer_type_detailed ~ ., data = bc_train.oversampled, 
                method = 'svmLinear3')

# Predict test data
bc_svm.pred <- predict(bc_svm, bc_test)
```

I created a simple Support Vector Machine model using the svmLinear3 parameter.

```{r}
# Create CrossTable
svm.cross <- CrossTable(x = bc_test$cancer_type_detailed, y = bc_svm.pred,
                        prop.chisq = F)

# Find accuracy
# accuracy.svm <- findAccuracy(svm.cross$t)
# paste('The accuracy of the model is:', round(accuracy.svm, 4))

# Find kappa
kappa.svm <- Kappa(svm.cross$t)
paste('The Kappa Statistic is:', round(kappa.svm$Unweighted[[1]], 4))
```

Here we can see the accuracy of our model is 70%, which is less than the kNN
model. However, the kappa statistic has increased to 0.15.


## ANN

```{r neural-net include=FALSE}
# Train the model
bc_ann <- train(cancer_type_detailed ~ ., data = bc_train.oversampled, method = 'avNNet', 
                na.action = na.omit)

# Find predictions
bc_ann.pred <- predict(bc_ann, bc_test)
```

```{r evaluate-nn}
# Create CrossTable
ann.cross <- CrossTable(x = bc_test$cancer_type_detailed, y = bc_ann.pred,
                        prop.chisq = F)

# Find accuracy
# accuracy.ann <- findAccuracy(ann.cross$t)
# paste('The accuracy of the model is:', round(accuracy.ann, 4))

# Find kappa
kappa.ann <- Kappa(ann.cross$t)
paste('The Kappa Statistic is:', round(kappa.ann$Unweighted[[1]], 4))
```

Neural network has classified everything as the Breast Invasive Ductal Carcinoma.

